---
title: "Isolated sound change"
author: "Stefano Coretta"
date: "18/04/2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(zipfR)
library(ggplot2)
theme_set(theme_bw())
library(dplyr)
library(lme4)
library(afex)
library(effects)
library(mgcv)
library(itsadug)
```

# General functions

The frequency distribution is produced by the following function, `zipfDistr()`. The distribution is a random sample from a Zipfian distribution (it uses the `zipfR` package).

```{r zipfDistr}
zipfDistr <- function(size) {
    zipf.model <- lnre("zm", alpha = 2/7, B = 0.1)
    zipf.sample <- rlnre(zipf.model, n = size)
    zipf.distr <- sort(as.numeric(as.character(zipf.sample)))
    return(zipf.distr)
}
```

We then need a function for generating values for the BART and BAT vowels in the lexicon. Since the main change affected the second formant (F1), the function will return F1 values (in Bark). The function takes a `target` value, it applies random noise and a `bias` (if specified; the default is `0`). A noise window of 0.5 (between -0.25 and 0.25) is used for the application of the production noise.

```{r produceFormants}
produceFormants <- function(target, bias = 0) {
    noise <- sample(seq(-50, 50, by = 10), 1)
    outcome <- target + noise + bias
    return(outcome)
}
```

# /ʊ/-fronting simulation

## Functions

```{r createLexicon}
createLexicon <- function(size, frequency) {

    word <- seq(1, size)
    cloud <- list()

    lexicon <- list()

    for (i in 1:size) {
        lexicon[[i]] <- list(word[i], frequency[i], cloud)
    }

    return(lexicon)
}
```

```{r populateLexicon}
populateLexicon <- function(lexicon, iterations, frequency, f2) {
    for (i in 1:iterations) {
        item <- sample(seq(1, length(lexicon)), 1, prob = frequency)
        cloud.size <- length(lexicon[[item]][[3]])
            if (cloud.size == 0) {
                lexicon[[item]][[3]][[1]] <- produceFormants(f2)
            } else if (cloud.size == 1) {
                target <- lexicon[[item]][[3]][[1]]
                lexicon[[item]][[3]][[cloud.size + 1]] <- produceFormants(target)
            } else {
                target <- sample(unlist(lexicon[[item]][[3]]), 1)
                lexicon[[item]][[3]][[cloud.size + 1]] <- produceFormants(target)
            }
    }
    return(lexicon)
}
```

## Simulate change

```{r lexicon}
set.seed(888)

lexicon.size <- 1000
frequency <- zipfDistr(lexicon.size)
lexicon <- createLexicon(lexicon.size, frequency)

iterations <- 50000
f2 <- 1000
#lexicon <- populateLexicon(lexicon, iterations, frequency, f2)
source.lexicon <- lexicon

change <- data.frame("word" = numeric(),
                     "time" = numeric(),
                     "frequency" = numeric(),
                     "f1" = numeric()
                     )
```

```{r u-fronting}
iterations <- 200000
bias <- 100

for (iteration in 1:iterations) {
    word <- sample(1:lexicon.size, 1, prob = frequency)

    #### Produce the chosen word ####

    if (length(lexicon[[word]][[3]]) == 0) {
        lexicon[[word]][[3]] <- list(produceFormants(f2, bias))
    } else {
        target <- sample(1:length(lexicon[[word]][[3]]), 1)
        outcome <- produceFormants(lexicon[[word]][[3]][[target]], bias)
        lexicon[[word]][[3]][[length(lexicon[[word]][[3]]) + 1]] <- outcome
    }
    
    #### Categorisation process ####
    # encoding.prob <- lexicon[[word]][[2]] /
    #                 max(frequency)
    # encode <- sample(c("yes", "no"),
    #                  1,
    #                  prob = c(encoding.prob,
    #                           1 - encoding.prob
    #                  )
    # )
    # 
    # #### Encoding function ####
    # if (encode == "no") {
    #         if (length(lexicon[[word]][[3]]) != 0) {
    #             for (j in length(lexicon[[word]][[3]]):1) {
    #                 if (lexicon[[word]][[3]][[j]] == outcome) {
    #                     if (length(lexicon[[word]][[3]]) == 1) {
    #                         lexicon[[word]][[3]] <- list()
    #                     } else {
    #                         lexicon[[word]][[3]][[j]] <- NULL
    #                     }
    #                 }
    #             }
    #         }
    # }


    if (iteration %% 1000 == 0) {
        current.frequency <- NULL
        f2.mean <- NULL
        for (word in 1:lexicon.size) {
            word.frequency <- lexicon[[word]][[2]]
            current.frequency <- c(current.frequency, word.frequency)
            if (is.null(unlist(lexicon[[word]][[3]])) == FALSE) {
                word.f2.mean <- mean(unlist(lexicon[[word]][[3]]))
            } else {
                word.f2.mean <- NA
            }
            f2.mean <- c(f2.mean, word.f2.mean)
        }
        words <- seq(1, lexicon.size)
        time <- rep(iteration, lexicon.size)
        current.change <- data.frame("word" = words, time, "frequency" = current.frequency, "f2" = f2.mean)
        change <- rbind(change, current.change)
    }
}
```

## Plotting

```{r freq-bin}
change.bin <- mutate(change,
                 freq.bin = ifelse(frequency < max(frequency)/2,
                                   "low",
                                   "high"
                                   )
                 )
```

```{r freq-plot}
change.bin %>%
ggplot(aes(time, f2, colour = freq.bin)) +
    geom_smooth(se = FALSE) +
    xlab("time (iterations)") +
    ylab("F2 (Hertz)") +
    scale_colour_discrete(name = "Frequency bin")
```

## LMER

```{r lmer}
change.scaled <- change %>%
    mutate(frequency = scale(frequency),
           time = scale(time))

u.fronting.lmer <- lmer(
    f2 ~
        time *
        frequency +
        (1|word),
    data = change.scaled,
    REML = FALSE
)

summary(u.fronting.lmer)

u.fronting.lmer.null <- lmer(
    f2 ~
        time +
        frequency +
        (1|word),
    data = change.scaled,
    REML = FALSE
)

anova(u.fronting.lmer.null, u.fronting.lmer)
```

## GAMM

```{r change-ordered}
change$word.ord <- as.ordered(change$word)
contrasts(change$word.ord) <- "contr.treatment"
change <- change %>%
    na.omit() %>%
    arrange(word, time) %>%
    start_event(column = "time", event = c("word"))
```

```{r gamm}
u.fronting.gamm <- bam(
    f2 ~
        frequency +
        s(time, bs = "cr") +
        s(time, by = frequency, bs = "cr"),
    data = change,
    method = "ML"
)

summary(u.fronting.gamm)

u.fronting.gamm.null <- bam(
    f2 ~
        frequency +
        s(time, bs = "cr"),
    data = change,
    method = "ML"
)

compareML(u.fronting.gamm.null, u.fronting.gamm)
```

```{r gamm-acr}
acf_plot(resid(u.fronting.gamm), split_by=list(change$word.ord))
```

```{r gamm-ar}
rho <- start_value_rho(u.fronting.gamm)

u.fronting.gamm <- bam(
    f2 ~
        frequency +
        s(time, bs = "cr") +
        s(time, by = frequency, bs = "cr"),
    data = change,
    method = "fREML",
    rho = rho,
    AR.start = change$start.event
)

summary(u.fronting.gamm)

u.fronting.gamm.null <- bam(
    f2 ~
        frequency +
        s(time, bs = "cr"),
    data = change,
    method = "fREML",
    rho = rho,
    AR.start = change$start.event
)

compareML(u.fronting.gamm.null, u.fronting.gamm)
```

```{r gamm-ar-acr}
acf_resid(u.fronting.gamm.null, split_pred="AR.start")
```

```{r simulation2-gam-ar-plot}
fvisgam(u.fronting.gamm.null, view = c("time", "frequency"),
        ylim = quantile(change$frequency, c(0, 1))
        )
```


